---
title: "Analysis_Final_Smruthi"
author: "Smruthi"
date: "2024-06-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(ggplot2)
```

## Load Datasets
Load the datasets from the processed folder.
```{r}
file_budget <- "data/processed/la_county_budget.csv"
file_employees <- "data/processed/la_employees.csv"

budget <- read.csv(file_budget)
employees <- read.csv(file_employees)
```

## Look at how number of employees in each ethnicity group has changed over years
```{r}
employees_only <- read.csv(file_employees)
employees_only <- employees_only %>% filter(QUARTER == 4)
employees_only <- employees_only %>%
  mutate(ETHNICITY_DESC = case_when(
    ETHNICITY_DESC %in% c('AMERICAN INDIAN OR ALASKA NATIVE', 'TWO or MORE RACES', 'NATIVE HAWAIIAN or OTHER PACIFIC ISLANDER','AMERICAN INDIAN','DEFAULT - CONVERSION') ~ 'OTHER',
    ETHNICITY_DESC == 'BLACK OR AFRICAN AMERICAN' ~ 'BLACK/AA',
    ETHNICITY_DESC == 'HISPANIC OR LATINO' ~ 'HISP/LAT',
    ETHNICITY_DESC == 'HISPANIC' ~ 'HISP/LAT',
    ETHNICITY_DESC == 'BLACK' ~ 'BLACK/AA',
    ETHNICITY_DESC == 'ASIAN / PACIFIC ISLANDERS' ~ 'ASIAN',
    TRUE ~ ETHNICITY_DESC
  ))
employee_count_yr <- employees_only %>%
  group_by(YEAR,ETHNICITY_DESC) %>%
  summarise(TOTAL_EMPLOYEE_COUNT = sum(EMPLOYEE_COUNT))%>%
  arrange(desc(TOTAL_EMPLOYEE_COUNT))

png(
file.path("Employee_Count_vs_Ethnicity_Years.png"),
width=1080,
height=800,
pointsize=30
)
ggplot(employee_count_yr, aes(x = YEAR, y = TOTAL_EMPLOYEE_COUNT, color = ETHNICITY_DESC, group = ETHNICITY_DESC)) +
  geom_line(size = 1) +  # Line for each ethnicity
  geom_point(size = 2) + # Points for each year
  geom_text(data = subset(employee_count_yr, YEAR %in% c(2015, 2022)),
            aes(label = TOTAL_EMPLOYEE_COUNT), vjust = -0.5, size = 3) +
  labs(title = "Employee Count by Ethnicity Across Years",
       x = "Year",
       y = "Employee Count",
       color = "Ethnicity") +
  scale_x_continuous(breaks = unique(employee_count_yr$YEAR)) +  
  scale_y_continuous(breaks = seq(0, max(employee_count_yr$TOTAL_EMPLOYEE_COUNT), by = 5000)) +
  theme_minimal() +  # Clean theme
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
dev.off()
```
Take a look at the first few rows of each dataset and the unique values in the year columns.
```{r}
#head(budget)
#head(employees)

# Show unique years in budget
unique(budget$Budget.Fiscal.Year)

# Show unique year and quarter in employees
unique(employees[, c("YEAR", "QUARTER")])
```
Filter the data to only include the years in common
```{r}
years <- intersect(unique(budget$Budget.Fiscal.Year), unique(employees$YEAR))

budget <- budget %>% filter(Budget.Fiscal.Year %in% years)
employees <- employees %>% filter(YEAR %in% years)
```

Checking for number of departments and job descriptions in employees table
```{r}
unique_departments <- employees %>%
  distinct(DEPARTMENT_NAME)
unique_departments_count <- employees %>%
  summarise(UNIQUE_DEPARTMENTS = n_distinct(DEPARTMENT_NAME))
unique_positions <- employees %>%
  distinct(EEOF_LONG_DD)
```

For employees, we will only consider the employees at the end of the year.
```{r}
 employees <- employees %>% filter(QUARTER == 4)
```

## Ethnicity breakout of employees data
```{r}
# Cleaning up Ethnicity_desc variable
employees1 <- employees %>%
  mutate(ETHNICITY_DESC = case_when(
    ETHNICITY_DESC %in% c('AMERICAN INDIAN OR ALASKA NATIVE', 'TWO or MORE RACES', 'NATIVE HAWAIIAN or OTHER PACIFIC ISLANDER') ~ 'OTHER',
    ETHNICITY_DESC == 'BLACK OR AFRICAN AMERICAN' ~ 'BLACK/AA',
    ETHNICITY_DESC == 'HISPANIC OR LATINO' ~ 'HISP/LAT',
    TRUE ~ ETHNICITY_DESC
  ))

# Removing GENDER_DESC = NonBinary only 3 observations
employees1 <- employees1 %>%
  filter(GENDER_DESC != "NonBinry")

ethnicity_breakout <- employees1 %>%
  group_by(ETHNICITY_DESC) %>%
  summarise(TOTAL_EMPLOYEE_COUNT = sum(EMPLOYEE_COUNT))

ethnicity_breakout<- ethnicity_breakout %>%
  mutate(Total_Employee_Count_Thousands = TOTAL_EMPLOYEE_COUNT / 1e3)

ggplot(ethnicity_breakout, aes(x = reorder(ETHNICITY_DESC, -Total_Employee_Count_Thousands), y = Total_Employee_Count_Thousands)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Employee Counts by Ethnicity",
       x = "Ethnicity",
       y = "Total Employee Count(In Thousands)")
```

```{r}
png(
file.path("Employee_Count_vs_Ethnicity.png"),
width=1080,
height=800,
pointsize=30
)
ggplot(ethnicity_breakout, aes(x = reorder(ETHNICITY_DESC, -Total_Employee_Count_Thousands), y = Total_Employee_Count_Thousands)) +
  geom_bar(stat = "identity", position = position_dodge() , fill = "#33DD44AA") +
  labs(title = "Employee Counts by Ethnicity",
       x = "Ethnicity",
       y = "Total Employee Count(In Thousands)") +
  theme(plot.title = element_text(size = 30, hjust = 0.5),
axis.title.x = element_text(size = 30), # Adjust x-axis label size
axis.title.y = element_text(size = 30),
axis.text.y = element_text(size = 30),
axis.text.x = element_text(size = 30),
panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
dev.off()
```

Regrouping EEOF_LONG_DD(Department Long Description)
```{r}
func_code_map <- c(
  "FINANCIAL ADMIN GENERAL CONTROL" = "FIN ADMIN",
  "OTHER (SUPPORT SERVICES, LIBRARY AND ED)" = "OTHER",
  "NATURAL PARKS AND RECREATION" = "OTHER",
  "PUBLIC WELFARE" = "PUBLIC WELFARE",
  "FIRE PROTECTION" = "FIRE",
  "HOSPITALS AND SANITARIUMS" = "HOSPITALS",
  "HEALTH" = "HEALTH",
  "CORRECTIONS" = "CORRECTIONS",
  "POLICE PROTECTION" = "POLICE",
  "COMMUNITY DEVELOPMENT" = "OTHER",
  "UTILITIES AND TRANSPORTATION" = "UTIL & TRANS"
)

# Replace long descriptions with shorter labels
employees1$EEOF_LONG_DD <- func_code_map[employees1$EEOF_LONG_DD]
```

## Chi square test on Ethinicity_desc, Gender, Job Positions
```{r}
# Contingency table of ethnicity and gender
contingency_table <- table(employees1$ETHNICITY_DESC, employees1$GENDER_DESC)
print(contingency_table)
# Perform chi-square test of independence
chi_square_test <- chisq.test(contingency_table)

# Print the results
print(chi_square_test)
```

```{r}
contingency_table2 <- table(employees1$EEOF_LONG_DD, employees1$ETHNICITY_DESC)
chi_square_test <- chisq.test(contingency_table2)
print(contingency_table2)
# Print the results
print(chi_square_test)
```
## Gender breakout across departments
```{r}
gender_job_breakout <- employees1 %>%
  group_by(GENDER_DESC, EEOF_LONG_DD) %>%
  summarize(TOTAL_EMPLOYEE_COUNT = sum(EMPLOYEE_COUNT)) %>%
  ungroup()

gender_job_breakout <- gender_job_breakout %>%
  group_by(EEOF_LONG_DD) %>%
  mutate(percent = TOTAL_EMPLOYEE_COUNT / sum(TOTAL_EMPLOYEE_COUNT) * 100)

ggplot(gender_job_breakout, aes(x = EEOF_LONG_DD, y = TOTAL_EMPLOYEE_COUNT, fill = GENDER_DESC)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(percent, 1), "%")), position = position_stack(vjust = 0.5), size = 3) +
  labs(title = "Employee Counts by Gender and Department",
       x = "Department Group",
       y = "Total Employee Count",
       fill = "Gender") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
png(
file.path("Employee_Count_Gender_vs_Dept.png"),
width=1080,
height=800,
pointsize=30
)
ggplot(gender_job_breakout, aes(x = EEOF_LONG_DD, y = TOTAL_EMPLOYEE_COUNT, fill = GENDER_DESC)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(percent, 1), "%")), position = position_stack(vjust = 0.5), size = 6, fontface = "bold") +
  labs(title = "Employee Counts by Gender and Department",
       x = "Department Group",
       y = "Total Employee Count",
       fill = "Gender") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 15), 
axis.title.x = element_text(size = 25),
axis.title.y = element_text(size = 25),
axis.text.y = element_text(size = 25),
plot.title = element_text(size = 25),
legend.text = element_text(size = 15),
panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
dev.off()
```

```{r}
# Aggregate the data by ETHNICITY_DESC and EEO_FUNC_CD
ethnicity_job_breakout <- employees1 %>%
  group_by(ETHNICITY_DESC, EEOF_LONG_DD) %>%
  summarize(TOTAL_EMPLOYEE_COUNT = sum(EMPLOYEE_COUNT)) %>%
  ungroup()

ggplot(ethnicity_job_breakout, aes(x = EEOF_LONG_DD, y = TOTAL_EMPLOYEE_COUNT, fill = ETHNICITY_DESC)) +
  geom_bar(stat = "identity") +
  labs(title = "Employee Counts by Functional Code and Ethnicity",
       x = "Functional Code",
       y = "Total Employee Count",
       fill = "Ethnicity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


library(viridis)
ggplot(ethnicity_job_breakout, aes(x = EEOF_LONG_DD, y = ETHNICITY_DESC, fill = TOTAL_EMPLOYEE_COUNT)) +
  geom_tile() +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Employee Distribution Heatmap",
       x = "Job Description Category",
       y = "Ethnicity",
       fill = "Employee Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Jobs taken up by different ethnicity groups
```{r}
employees1$EEOF_JOB_CATEGORY_DESC <- toupper(employees1$EEO_JOB_CATEGORY_DESC)

job_ethnicity_s_breakout <- employees1 %>%
  group_by(ETHNICITY_DESC, EEOF_JOB_CATEGORY_DESC) %>%
  summarize(TOTAL_EMPLOYEE_COUNT = sum(EMPLOYEE_COUNT)) %>%
  filter(EEOF_JOB_CATEGORY_DESC %in% 
           c("SUPERVISING OPERATIVES", "SUPERVISING CRAFTS", "SUPERVISING GENERAL SERVICES", "SUPERVISING TECHNICAL", "SUPERVISING PARAPROFESSIONAL","SUPERVISING ADMINSTRATIVE STAFF", "SUPERVISING PROFESSIONAL", "SUPERVISING CLERICAL", "SUPERVISING PROTECTIVE & REGULATORY")) %>%
  ungroup()
```
```{r}
library(viridis)

job_ethnicity_s_breakout <- job_ethnicity_s_breakout %>%
  group_by(EEOF_JOB_CATEGORY_DESC) %>%
  mutate(PERCENTAGE = TOTAL_EMPLOYEE_COUNT / sum(TOTAL_EMPLOYEE_COUNT) * 100)

heatmap1 <- job_ethnicity_s_breakout %>%
  ggplot(aes(x = EEOF_JOB_CATEGORY_DESC, y = ETHNICITY_DESC, fill = PERCENTAGE)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Percentage") +
  geom_text(aes(label = sprintf("%.1f%%", PERCENTAGE)), color = "black", fontface = "bold") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1,size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 15),  # Adjust size of x-axis title
        axis.title.y = element_text(size = 15),
        plot.title = element_text(size = 15)) +
  labs(title = "Percentage of Employees by Ethnicity and Supervising Roles",
       x = "Job Category",
       y = "Ethnicity")

ggsave("heatmap.png", plot = heatmap1, width = 10, height = 8, dpi = 300, type = "cairo", bg = "white")
```


```{r}
job_ethnicity_s_breakout2 <- employees1 %>%
  group_by(ETHNICITY_DESC, EEOF_JOB_CATEGORY_DESC) %>%
  summarize(TOTAL_EMPLOYEE_COUNT = sum(EMPLOYEE_COUNT)) %>%
  filter(EEOF_JOB_CATEGORY_DESC %in% 
           c("MANAGEMENT", "EXECUTIVE MANAGEMENT", "MANAGEMENT STAFF")) %>%
  ungroup()

job_ethnicity_s_breakout2 <- job_ethnicity_s_breakout2 %>%
  group_by(EEOF_JOB_CATEGORY_DESC) %>%
  mutate(PERCENTAGE = TOTAL_EMPLOYEE_COUNT / sum(TOTAL_EMPLOYEE_COUNT) * 100)

heatmap2 <- job_ethnicity_s_breakout2 %>%
  ggplot(aes(x = EEOF_JOB_CATEGORY_DESC, y = ETHNICITY_DESC, fill = PERCENTAGE)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Percentage") +
  geom_text(aes(label = sprintf("%.1f%%", PERCENTAGE)), color = "black", fontface = "bold") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1,size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 15),  # Adjust size of x-axis title
        axis.title.y = element_text(size = 15),
        plot.title = element_text(size = 15)) +
  labs(title = "Percentage of Employees by Ethnicity and Management Role",
       x = "Job Category",
       y = "Ethnicity")

ggsave("heatmap2.png", plot = heatmap2, width = 10, height = 8, dpi = 300, type = "cairo", bg = "white")
```
```{r}
job_ethnicity_s_breakout3 <- employees1 %>%
  group_by(ETHNICITY_DESC, EEOF_JOB_CATEGORY_DESC) %>%
  summarize(TOTAL_EMPLOYEE_COUNT = sum(EMPLOYEE_COUNT)) %>%
  filter(EEOF_JOB_CATEGORY_DESC %in% 
           c("CRAFTS", "OPERATIVES", "TECHNICAL", "GENERAL SERVICE", "PARAPROFESSIONAL", "PROTECTIVE & REGULATORY", "PROFESSIONAL", "CLERICAL")) %>%
  ungroup()

job_ethnicity_s_breakout3 <- job_ethnicity_s_breakout3 %>%
  group_by(EEOF_JOB_CATEGORY_DESC) %>%
  mutate(PERCENTAGE = TOTAL_EMPLOYEE_COUNT / sum(TOTAL_EMPLOYEE_COUNT) * 100)

# Create the heatmap
heatmap3 <- job_ethnicity_s_breakout3 %>%
  ggplot(aes(x = EEOF_JOB_CATEGORY_DESC, y = ETHNICITY_DESC, fill = PERCENTAGE)) +
  geom_tile() +
  scale_fill_viridis_c(name = "Percentage") +
  geom_text(aes(label = sprintf("%.1f%%", PERCENTAGE)), color = "white", fontface = "bold") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Employee Percentage Across Ethnicities and Regular Jobs")

ggsave("heatmap3.png", plot = heatmap3, width = 10, height = 8, dpi = 300, type = "cairo", bg = "white")
```

```{r}
contingency_table3 <- table(employees1$EEOF_JOB_CATEGORY_DESC, employees1$ETHNICITY_DESC)
chi_square_test <- chisq.test(contingency_table3)
print(contingency_table3)
# Print the results
print(chi_square_test)
```
## Merging budget data with ethnicity data
```{r}
library(tidyverse)
employee_count_qtr_yr <- employees1 %>%
  group_by(EEOF_LONG_DD, ETHNICITY_DESC) %>%
  summarise(TOTAL_EMPLOYEE_COUNT = sum(EMPLOYEE_COUNT))%>%
  spread(key = ETHNICITY_DESC, value = TOTAL_EMPLOYEE_COUNT, fill = 0)

employee_count_qtr_yr1 <- employees1 %>%
  group_by(DEPARTMENT_CD, ETHNICITY_DESC) %>%
  summarise(TOTAL_EMPLOYEE_COUNT = sum(EMPLOYEE_COUNT)) %>%
  spread(key = ETHNICITY_DESC, value = TOTAL_EMPLOYEE_COUNT, fill = 0)

budget <- budget %>% filter(Budget.Fiscal.Year == 2022)

budget_by_year_department <- budget %>%
  group_by(Department.Code) %>%
  summarise(Total_Budget = sum(Adjusted.Budget)) %>%
  arrange(desc(Total_Budget))

data1 <- merge(
    budget_by_year_department,
    employee_count_qtr_yr1,
    by.x = c("Department.Code"),
    by.y = c("DEPARTMENT_CD")
)

dept_mapping <- data.frame(
  Department.Code = c("CS", "RP", "PB", "AD", "AN", "AO", "AS", "AU", "AW", "BS", 
               "CC", "CD", "DA", "HM", "IO", "IS", "PD", "RR", "TT", "FR", 
               "HA", "HC", "HJ", "HP", "HS", "MH", "PG", "PH", "PP", 
               "HG", "HH", "HN", "HO", "HR", "PR", "PS", "AR", "BH", "NH", 
               "PK", "RT", "AG", "CA", "CR", "EW", "JC", "MV", "PL", "YD", 
               "ME", "SH", "CH", "SS", "PW"),
  Category = c("OTHER", "OTHER", "CORRECTIONS", 
               "FIN ADMIN", "FIN ADMIN", "FIN ADMIN", "FIN ADMIN", "FIN ADMIN", 
               "FIN ADMIN", "FIN ADMIN", "FIN ADMIN", "FIN ADMIN", "FIN ADMIN", 
               "FIN ADMIN", "FIN ADMIN", "FIN ADMIN", "FIN ADMIN", "FIN ADMIN", 
               "FIN ADMIN", "FIRE", "HEALTH", "HEALTH", "HEALTH", 
               "HEALTH", "HEALTH", "HEALTH", "HEALTH", "HEALTH", "HEALTH", 
               "HOSPITALS", "HOSPITALS", "HOSPITALS", "HOSPITALS", "HOSPITALS", 
               "HOSPITALS", "HOSPITALS", "OTHER", "OTHER", 
               "OTHER", "OTHER", "OTHER", "OTHER", "OTHER", 
               "OTHER", "OTHER", "OTHER", "OTHER", "OTHER", "OTHER", 
               "POLICE", "POLICE", "PUB WELFARE", "PUB WELFARE", 
               "UTIL & TRANS")
)
merged_df <- merge(budget_by_year_department, dept_mapping, by = "Department.Code", all.x = TRUE)
data2 <- merge(
    merged_df,
    employee_count_qtr_yr,
    by.x = c("Category"),
    by.y = c("EEOF_LONG_DD")
)
# data1$Budget_by_Employee <- data1$Total_Budget / data1$TOTAL_EMPLOYEE_COUNT
```
## Run k-means clustering to see if certain ethnic groups belong to certain departments
```{r}
  # Remove Department.Code column
rownames(data1) <- data1$Department.Code  # Set row names as Department.Code
data_c <- data1[, -1]
data_c <- data_c[ , !(names(data_c) %in% c("Total_Budget"))]
# Standardize the data
scaled_data <- scale(data_c)

# Determine the optimal number of clusters using the elbow method
wss <- sapply(1:10, function(k) kmeans(scaled_data, k)$tot.withinss)
plot(1:10, wss, type = "b", xlab = "Number of Clusters", ylab = "Within-cluster sum of squares")

# Based on the plot, select the number of clusters (elbow point)
# Let's say we choose k = 5

# Perform k-means clustering
k <- 5
kmeans_result <- kmeans(scaled_data, centers = k)

# Add cluster labels to the original data
data_c$Cluster <- as.factor(kmeans_result$cluster)

# Print the cluster centers
print(kmeans_result$centers)

# View the cluster assignments
employees_data_with_clusters <- cbind(data1["Department.Code"], Cluster = kmeans_result$cluster)

# View the updated data frame
print(employees_data_with_clusters)
```





